Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-21 paper/ms.tex
update ms

This work is focused on Kepler stars for which ages can be inferred
using either rotation, lithium, or both.  These ``age-dateable stars''
are a minority of the \nkeplerstars\ Kepler targets.  Rotation periods
have been reported for roughly one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra
suitable for measurement of the \ion{Li}{1} 6708\,\AA\ doublet have
only been acquired for the Kepler objects of interest (KOIs), which
comprise a few percent of Kepler's targets.  In the following, we will
describe the set of stars that we adopt with measured rotation periods
(Section~\ref{subsec:rotsel}), the set of objects we adopt as hosting planets
(Section~\ref{subsec:planetsel}), and the subset of these with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).  Figure~\ref{fig:stellarprops}
provides a visualization of the various samples.

@@

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.  This
is not surprising, in that the rotation periods of KOIs have received
considerably more scrutiny than those of ordinary Kepler stars.  We
therefore decided to split our subsequent analysis into a homogeneous
portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.  For the
latter portion, we first included 32 KOIs with orbital and rotation
periods within $\approx$20\% that had been
excluded %\footnote{Excluding stars in this category would impose a bias
%against young close-in planets, due to the natural commensurability
%between the rotation rates of young stars, and Kepler's increased
%sensitivity to short-period objects.  Although some of these KOIs are
%known false positives, our analysis excludes such objects at later
%steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the \citetalias{Santos_2019} and \citetalias{Santos_2021}
catalogs (A.~Santos, private communication).  We then incorporated an
additional \nnewdavidtwentyone\ rotation periods for KOI hosts that
\citet{David_2021} had described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
and \citet{Angus_2018}.  Inclusion of these additional KOI rotation
periods is a supplementary measure aimed at completeness in our final
KOI age catalog; the provenances of the individual adopted periods are
noted in the relevant tables.

Finally, since our scope is focused on rotation-based age dating, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  FGK stars with slower rotation periods spin much
slower than the stars in the oldest open clusters currently used to
calibrate our gyrochronology model.  Figure~\ref{fig:prot_vs_teff}
shows the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.

@@

The final piece of our analysis involves assessing ages based on the
\ion{Li}{1} 6708\,\AA\ doublet.  Specifically, we analyzed spectra
from the High Resolution Echelle Spectrometer (HIRES;
\citealt{vogt_hires_1994}) on the Keck I 10m telescope.  These spectra
were primarily collected through the California Kepler Survey
\citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
We supplemented the existing archive with $\approx$10~hours of new
observations for 22 stars between Fall 2022 and Spring 2024.  These
stars were chosen to ensure that confirmed planets with rotational
evidence for ages below 1\,Gyr had spectra, since this is the age
range in which lithium is most likely to yield useful age constraints.

% stat from _estimate_new_lithium_timefrac.py
Lithium equivalent widths and abundances for the Kepler Objects of
Interest were already analyzed by \citet{2018ApJ...855..115B} for
roughly three quarters of the spectra in our sample.  However, new
spectra have since been acquired, and our approach and selection
function are different.  We therefore perform our own line width
measurements on
the reduced HIRES spectra.

We collected all blaze-corrected HIRES spectra from our group's
observations of non-false positive Kepler Objects of Interest with a
``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}) of at
least 10.  This yielded at least one spectrum for \nstwithspec\
stars hosting \nplwithspec\ planets.  About half of these stars have
rotation periods (\nstwithspecandprot\ stars and
\nplwithspecandprot, respectively).   For stars with multiple spectra
available at varying ranges of signal-to-noise (S/N), we analyzed only
the spectrum with the highest number of counts.  The resulting spectra
were acquired between 6 September 2009 and 16 May 2024, and
cumulatively comprise \nhireshours\,hours of open-shutter time.

We measured the lithium equivalent widths using a procedure adapted
from previous work \citep{Bouma_2021}.  Our stars of interest are FGK
stars, and so the continuum in the vicinity of the \ion{Li}{1}
6708\,\AA\ doublet is well-defined.  We Doppler corrected the spectra
to a common reference wavelength by cross-correlating against a high
S/N template for Kepler-1698, chosen because $|\gamma|$$<$10\kms,
$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
puts it in the middle of our sample's temperature range and gives it
mild line-broadening.  We then trimmed the Doppler-corrected spectra
to a local window centered on the lithium line, using a window width
of 15\,\AA\ (we also considered 10\,\AA\ and 20\,\AA; the results were
consistent).  We continuum-normalized by fitting a third-degree
Chebyshev series, while excluding regions with absorption lines.  We
then numerically integrated the resulting spectrum using a
one-component Gaussian with free amplitude, width, and mean, and
estimated uncertainties on the line width through a Monte Carlo
procedure that bootstrapped against the local scatter in the continuum.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch %\footnote{\url{https://gaia-kepler.fun}, last
%accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
%because 91 KOIs are omitted in the 1$''$ match, 23 of which have
%either ``candidate'' or ``confirmed'' status.  Almost all of these 23
%are M dwarfs with high proper motions.  Performing the same crossmatch
%at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
%false positives, fail to yield Gaia DR3 crossmatches.}, 
%which is a
%crossmatch
of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The
separation distribution of the Kepler-Gaia DR3 crossmatches is such
that 99.2\% of candidate matches are within 1$''$.   We nonetheless
noted an upturn in the candidate match rate from 3-4$''$; such sources
are flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ radius, we adopted the brightest star as the
default match.  In most such cases this was unambiguous because there
is a large brightness difference between the primary and any apparent
neighbors.  However cases with multiple stars within 4$''$ within
$\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in both panels of
Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
ages because {\it i)} point sources far from the main sequence include
unresolved photometric binaries, and {\it ii}) effects other than
magnetic braking influence the rotation periods of bona fide single
stars far from the main sequence.  The added complications of
reddening and metallicity further complicate matters.  After exploring
various options, we settled on the orange locus in the $\log
g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
the details of how exactly this locus is constructed are arbitrary, 
%and convoluted,\footnote{In practice, for K dwarfs we fitted an
%$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
%B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
%$t_{\rm B20,iso}$ was the isochronal age reported by
%\citet{Berger_2020a_catalog}. We let the order of the fit vary from
%$N$=1 to 10, minimized the Bayesian Information Criterion,
%and found $N$=6.  For F and G dwarfs, this procedure failed to produce
%a reasonable locus because of shifts in the relative precision of
%the isochrone ages.  We therefore visually selected KOIs that appeared
%to be near the ``main sequence'', fitted a separate polynomial to
%their locations, and then stitched the resulting polynomials from the
%F and G dwarfs to that of the K dwarfs.}
the general aim is to flag stars for which there is evidence based on
their location in the $\log g$--$T_{\rm eff}$ plane that our
gyrochronology model may not be reliable.  
%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or
Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
the period distribution of this population differs from stars near the
main sequence.

@@

{\it Field star ages from other surveys}:
While this study focused on deriving rotation and lithium-based ages
for Kepler planets, surveys such as K2, TESS, HAT, WASP, NGTS, ZTF,
and ATLAS have different selection functions.  Applying rotation-based
age dating to their discoveries could clarify the youth and
evolutionary trends of these populations.  Future prospects like PLATO
\citep{Rauer14}, Earth 2.0 \citep{2022arXiv220606693G}, and the Roman
Galactic Plane Time Domain Survey seem poisted to further expand the
sample of stars with precisely determined ages, including those known
to host planets.

>>>>

This work is focused on Kepler stars for which ages can be inferred
using either rotation, lithium, or both.  Such stars
are a minority of the \nkeplerstars\ Kepler targets.  Rotation periods
have been reported for roughly one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra
suitable for measurement of the \ion{Li}{1} 6708\,\AA\ doublet have
only been acquired for the Kepler objects of interest (KOIs), which
comprise a few percent of Kepler's targets.  In the following, we will
describe the set of stars that we adopt with measured rotation periods
(Section~\ref{subsec:rotsel}), the set of objects we adopt as hosting planets
(Section~\ref{subsec:planetsel}), and the subset of these with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).  Figure~\ref{fig:stellarprops}
provides a visualization of the various samples.

@@

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.  This
is not surprising, in that the rotation periods of KOIs have received
considerably more scrutiny than those of ordinary Kepler stars.  We
therefore decided to split our subsequent analysis into a homogeneous
portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.  For the
latter portion, we first included 32 KOIs with orbital and rotation
periods within $\approx$20\% that had been
excluded %\footnote{Excluding stars in this category would impose a bias
%against young close-in planets, due to the natural commensurability
%between the rotation rates of young stars, and Kepler's increased
%sensitivity to short-period objects.  Although some of these KOIs are
%known false positives, our analysis excludes such objects at later
%steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the \citetalias{Santos_2019} and \citetalias{Santos_2021}
catalogs (A.~Santos, private communication).  We then incorporated an
additional \nnewdavidtwentyone\ rotation periods for KOI hosts that
\citet{David_2021} had described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013}, \citet{Walkowicz_2013}, \citet{Mazeh_2015}
and \citet{Angus_2018}.  Inclusion of these additional KOI rotation
periods is a supplementary measure aimed at completeness in our final
KOI age catalog; the provenances of the individual adopted periods are
noted in the relevant tables.

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK
stars in the open clusters used to
calibrate our gyrochronology model have $P_{\rm rot}$$\approx$35\,days.
Figure~\ref{fig:prot_vs_teff}
shows the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.

@@

The final piece of our analysis involves assessing ages based on the
\ion{Li}{1} 6708\,\AA\ doublet.  We analyzed spectra
from the High Resolution Echelle Spectrometer (HIRES;
\citealt{vogt_hires_1994}) on the Keck I 10m telescope.  These spectra
were primarily collected through the California Kepler Survey
\citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
We supplemented the existing archive with $\approx$10~hours of new
observations for 22 stars between Fall 2022 and Spring 2024.  These
stars were chosen to ensure that confirmed planets with rotational
evidence for ages below 1\,Gyr had spectra, since this is the age
range in which lithium is most likely to yield useful age constraints.

% stat from _estimate_new_lithium_timefrac.py
Lithium equivalent widths and abundances for the Kepler Objects of
Interest were already analyzed by \citet{2018ApJ...855..115B} for
roughly three quarters of the spectra in our sample.  However, new
spectra have since been acquired, and our approach and selection
function are different.  We therefore performed our own line width
measurements on
the reduced HIRES spectra.

We collected all blaze-corrected HIRES spectra from our group's
observations of non-false positive Kepler Objects of Interest with a
``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}) of at
least 10.  This yielded at least one spectrum for \nstwithspec\
stars hosting \nplwithspec\ planets.  About half of these stars have
measured rotation periods (\nstwithspecandprot\ stars and
\nplwithspecandprot, respectively).   For stars with multiple spectra
available, we analyzed only
the spectrum with the highest number of counts.  The resulting spectra
were acquired between 6 September 2009 and 16 May 2024, and
cumulatively comprise \nhireshours\,hours of open-shutter time.

We measured the lithium equivalent widths using a procedure adapted
from previous work \citep{Bouma_2021}.  Our stars of interest are FGK
stars, and so the continuum in the vicinity of the \ion{Li}{1}
6708\,\AA\ doublet is well-defined.  We Doppler-corrected the spectra
to a common reference wavelength by cross-correlating against a high
S/N template for Kepler-1698, chosen because $|\gamma|$$<$10\kms,
$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
puts it in the middle of our sample's temperature range and gives it
mild line-broadening.  We then trimmed the Doppler-corrected spectra
to a local window centered on the lithium line, using a window width
of 15\,\AA\ (we also considered 10\,\AA\ and 20\,\AA; the results were
consistent).  We continuum-normalized by fitting a third-degree
Chebyshev series, while excluding regions with absorption lines.  We
then numerically integrated the resulting spectrum using a
one-component Gaussian with free amplitude, width, and mean, and
estimated uncertainties on the line width through a Monte Carlo
procedure that bootstrapped against the local scatter in the continuum.

@@

{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
Kepler-to-Gaia crossmatch %\footnote{\url{https://gaia-kepler.fun}, last
%accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
%because 91 KOIs are omitted in the 1$''$ match, 23 of which have
%either ``candidate'' or ``confirmed'' status.  Almost all of these 23
%are M dwarfs with high proper motions.  Performing the same crossmatch
%at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
%false positives, fail to yield Gaia DR3 crossmatches.}, 
%which is a
%crossmatch
of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3 (accessible
online at \url{https://gaia-kepler.fun}).  The
separation distribution of the Kepler-Gaia DR3 crossmatches is such
that 99.2\% of candidate matches are within 1$''$.   We nonetheless
noted an upturn in the candidate match rate from 3-4$''$; such sources
are flagged using bit 4.  For KIC stars with multiple potential Gaia
matches within the 4$''$ radius, we adopted the brightest star as the
default match.  In most such cases this was unambiguous because there
is a large brightness difference between the primary and any apparent
neighbors.  However cases with multiple stars within 4$''$ within
$\Delta G$$<$$0.5$\,mag are noted using bit flag 5.

@@

{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in both panels of
Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
ages because {\it i)} point sources far from the main sequence include
unresolved photometric binaries, and {\it ii}) effects other than
magnetic braking influence the rotation periods of bona fide single
stars far from the main sequence.  The added concerns of
reddening and metallicity further complicate matters.  After exploring
various options, we settled on the orange locus in the $\log
g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
the exact details of how this locus is constructed are arbitrary\footnote{For purposes of reproducibility, 
	the method was as follows.
The orange line in Figure~\ref{fig:stellarprops} is	defined as
the maximum of two numerically-determined functions, $f_0(T_{\rm eff})$ and $f_1(T_{\rm eff})$.
We defined $f_0$ by fitting an
$N^{\rm th}$-order polynomial to the stars with rotation periods for which $t_{\rm
B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$
and $T_{\rm eff}\in[3700,5900]$\,K, where
$t_{\rm B20,iso}$ was the isochronal age reported by
\citet{Berger_2020a_catalog}.  We let the order of the fit vary from
$N$=1 to 10, minimized the Bayesian Information Criterion,
and found $N$=6.  
By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
However for F and G dwarfs, the resulting locus
allowed for stars that were `too evolved''; the density of KOIs 
was anomalously low in this region of the $\log g$ vs.\ $T$$_{\rm eff}$ plane.
We therefore visually selected KOIs that appeared
to be near the main sequence -- i.e.~ most of the blue
points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
polynomial $f_1$ through a similar BIC-minimization procedure,
and found $N$=3.
The portion of the orange locus from $\approx$5300--6200\,K
is set as $f_1 + c$, for $c$ a constant offset that we visually tuned to be
$\approx$0.1\,dex above the ``KOI main sequence''.
The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
We emphasize that for stars
surface gravities above the orange locus, we do report rotation based ages;
they are simply flagged as not being near the main sequence.  }
the general aim is to flag stars for which there is evidence based on
their location in the $\log g$--$T_{\rm eff}$ plane that our
gyrochronology model may not be reliable.  
%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or
Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
the period distribution of this population differs from stars near the
main sequence.

@@

{\it Field star ages from other surveys}:
While this study focused on
Kepler, other photometric surveys (e.g.~K2, TESS, HAT, WASP, NGTS, ZTF,
ATLAS) open opportunities for age-dating a far broader set
of stars and planets.
Future prospects also include PLATO
\citep{Rauer14}, Earth 2.0 \citep{2022arXiv220606693G}, and the Roman
Galactic Plane Time Domain Survey.
The long-baseline surveys 
that specifically target different regions of the Galaxy
seem most likely to yield new insight into its star
formation history.
